Contracts for Haskell

### 2011-06-30 contracts.tex

\subsection{FOL}
\begin{center}
\begin{array}{rclr}
  t &:=& x \mid \mbox{app}(t_1,t_2) \mid D(t,\dots,t) \mid f \mid n \mid \bad \mid \unr \mid \mbox{CF}(t) & (Term)\\
  \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula)\\
\end{array}
\end{center}
We always give the following equations to define crash-freeness:
$$\cf{\unr}, \lnot \cf{\bad}, \forall f,x~\cf{f} \land \cf{x} \implies \cf{app(f,x)}$$
Note that it's only the equation we \textit{always} give, but there some other equations that will arise from the translations below.
It does not follow exactly the semantics given by the popl paper, but we hope it'll be a sufficient approximation!

\subsection{Contracts}
\begin{center}
\begin{array}{rclr}
  c &:=& x:c_1 \to c_2\\
  &\mid& (c_1,c_2) \\
  &\mid& \{ x \mid e \}\\
  &\mid& \any \\
\end{array}
\end{center}

Semantics of contract satisfaction (exactly the same as in the POPL paper):
\begin{center}
\begin{array}{rcl}
  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
  e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
  e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
  e \in \any &\iff& True  
\end{array}
\end{center}

@@

Equation (10) is not restrictive even in the case of laziness because $e$ is always evaluated. (TODO, but not always fully evaluated...)

@@

\subsection{$\strans{}$}
$\strans{e \in c}$ is a set of first-order formulae.
\begin{eqnarray}
\strans{e \in \any} &=& true\\
\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etranst{e}) \land \etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
&& \cup \etransf{e} \cup \etransf{u[e/x]}\\
\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
\end{eqnarray}
$False$ is a data constructor here.

@@

@@

copy ::: a:{x : True} -> {y:True};;
\end{verbatim}
The contract states that if we give to copy a crash-free argument $x$, copy will return a crash-free result $y$.
Note that [Just BAD] is not crash-free in our sense, so we can't say anything about it. TODO

@@

add ::: a:{x : isZero x} -> {y:isZero y} -> {z: isZero z};;
\end{verbatim}

@@

notZero ::: a:{x:True} -> {y: True};;
add  ::: a:{x: True} -> b:{y: True} -> {z: True};;
add  ::: a:{x: notZero x} -> b:{y: True} -> {z: True};;
mult ::: a:{x: notZero x} -> b:{y: notZero y} -> {z: True};;
\end{verbatim}

@@

\subsection{Removing CFness}
In the POPL paper, CF was introduced in the contract satisfaction
semantics because it was required in order to define the
wrapping. Given that we do not use any wrapping, we can remove it from
the semantics.

In fact, it's probably better to decouple crash-freeness and predicate
satisfaction.

Here the new version of the semantics : $e \in \{x \mid p \} \iff e
\mbox{ diverges or } (p[e/x] \not \to^\star \{\bad,False\})$

But it leads to some tricky cases:

A possible fix is to provide, directly in the source language a $cf$
predicate, so that one can write contract like:

$$add ::: a:\{ x \mid cf(x) \} \to \{ y \mid cf(y) \} \to \{ z \mid z
\geq a \}$$

Note that in this setting $\{ x \mid True \} = \any$ and $\{ x \mid
cf(x) \} = \ok$.

\subsection{Dimitrios point}
Dimitrios isn't convinced that we keep the consistency of the whole
module when checking one function after the other.

Assume that we have a module that contains two functions defintion $f$
and $g$ and two contracts : $f ::: Any -> Ok$ and $g ::: Any ->
Ok$. We assume that those contract do not hold, for example if $f$ is
$not$ and $g$ is $id$.

Finally, the user is really happy, but in fact he has proved nothing.

I think that it's not an actual issue, if we only give to equinox
useful formulae. Ie, if $f$ does not use $g$ we do not include $g$ in
the translation. Obviously it slows down the proof and, breaks the
consistency.

Consider the following graph, an arrow from $a$ to $b$ meaning that
$b$ uses $a$ in its definition.

We (D and I) believe that if we check a function by only giving the
formulae coming from the ancestors, then we do not break consistency.
Eg, to check $a$'s contract we need to give the contract and
definition for $f$ and $g$ too, but not for $h$.

>>>>

\subsection{FOL}
\begin{center}
\begin{array}{rclr}
  t &:=& x \mid \mbox{app}(t_1,t_2) \mid D(t,\dots,t) \mid f \mid n \mid \bad \mid \unr \mid \mbox{CF}(t) & (Term)\\
  \phi &:=& \forall x.\phi \mid \phi \to \phi \mid \lnot \phi \mid \phi \lor \phi \mid \phi \land \phi \mid true \mid t=t \mid \mbox{CF}(t)& (Formula)\\
\end{array}
\end{center}

We always give the following equations to define crash-freeness:
$$\cf{\unr}, \lnot \cf{\bad}, \forall f,x~\cf{f} \land \cf{x} \implies \cf{app(f,x)}$$
Note that it's only the equation we \textit{always} give, but there some other equations that will arise from the translations below.
It does not follow exactly the semantics given by the popl paper, but we hope it'll be a sufficient approximation!

\subsection{Contracts}
\begin{center}
\begin{array}{rclr}
  c &:=& x:c_1 \to c_2\\
  &\mid& (c_1,c_2) \\
  &\mid& \{ x \mid e \}\\
  &\mid& \{ !x \mid e \}\\
\end{array}
\end{center}

Semantics of contract satisfaction (exactly the same as in the POPL paper):
\begin{center}
\begin{array}{rcl}
  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
  e \in \{!x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
  e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
  e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
\end{array}
\end{center}

\section{CF-ness and unreachability}
\subsection{CF-ness}
In the POPL paper, CF was introduced in the contract satisfaction
semantics because it was required in order to define the
wrapping. Given that we do not use any wrapping, we can make a
separation between crash-freeness and predicate satisfaction now, as
shown in the contracts semantics, with the special syntax.

It leads to some tricky cases:

Now, the correct contract is:
$$add ::: a:\{ !x \mid cf(x) \} \to \{ !y \mid cf(y) \} \to \{ !z \mid z
\geq a \}$$

@@

Equation (10) is not restrictive even in the case of laziness because $e$ is always evaluated.

@@

\subsection{$\strans{}$}
$\strans{e \in c}$ is a set of first-order formulae.
\begin{eqnarray}
\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etranst{e}) \land \etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
&& \cup \etransf{e} \cup \etransf{u[e/x]}\\
\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
\end{eqnarray}
$False$ is a data constructor here.

@@

\subsection{Checking the contracts in the right order}
Here is a little example showing that we should be careful about we
belong to a theory.

Assume that we have a module that contains two functions defintion $f$
and $g$ and two contracts : $f ::: Any \to Ok$ and $g ::: Any \to
Ok$. We assume that those contract do not hold, for example if $f$ is
$not$ and $g$ is $id$.

Finally, the user is really happy, but in fact he has proven nothing.

The solution is to only include the formulae that belongs to a
functions that are actually used. For example, to prove $a$'s
contract, we should only include $f$ and $g$ formulae.

@@

copy ::: a:{!x : True} -> {!y:True};;
\end{verbatim}
The contract states that if we give to copy a crash-free argument $x$, copy will return a crash-free result $y$.
Note that [Just BAD] is not crash-free in our sense, so we can't say anything about it.

@@

add ::: a:{!x : isZero x} -> {!y:isZero y} -> {!z: isZero z};;
\end{verbatim}

@@

notZero ::: a:{!x:True} -> {!y: True};;
add  ::: a:{!x: True} -> b:{!y: True} -> {!z: True};;
add  ::: a:{!x: notZero x} -> b:{!y: True} -> {!z: True};;
mult ::: a:{!x: notZero x} -> b:{!y: notZero y} -> {!z: True};;
\end{verbatim}

@@
